---
title: "NYPD data"
author: "PWONG"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NYPD Shooting Incident Data
The objective of this exercise is to explore what the data can tell us.
The objective is kept broad instead of formulating questions to be answered by this limited set of data to avoid potential biases since it is a new set of data which is being looked at for the first time. 

### Load library & clear environment
```{r}
rm(list=ls())
library(tidyverse)
library(lubridate)
library(ggplot2)
```

### Get & read data on csv file
Upon looking at the data, it can be seen that **the data falls into 3 large categories of information**.

1. logistical information (date, time, location)
2. perpetrator information (age, sex, race)
3. victim information (age, sex, race, murder count)

Based on these information, **2 realizations** are made.

1. Upon cleaning up the data, it can give us a simplistic view of shooting incidents in New York.
  + the data needs to be cleaned up as some of the data needs to be converted to different data types to facilitate analysis
  + some cases with too much data marked as N/A may not be suitable to be used for analysis
2. Caution needs to be exercised in running any statistical models or coming to any conclusions based on just this data alone.
  + missing data in certain variables, e.g. data & information gathered about perpetrators only include those who could be identified (no info on those who are not caught)
  +	different factors other than personal characteristics (such as gun policies in the state) need to be considered to draw conclusions on shooting incidents otherwise it may perpetuate unnecessary biases

Based on the above, it was **decided to analyze the data about murder victims of shooting incidents across the years in different locations in New York**.
```{r get_data}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD_data <- read_csv(url)
summary(NYPD_data)
```

### Clean & select NYPD data
```{r clean_data}
## convert to date and factor
## create new column "year"
## select only necessary columns
NYPD_data <- NYPD_data %>%
  mutate(OCCUR_DATE=mdy(OCCUR_DATE), 
         BORO=as.factor(BORO),
         PRECINCT=as.factor(PRECINCT),
         VIC_AGE_GROUP=as.factor(VIC_AGE_GROUP),
         VIC_RACE=as.factor(VIC_RACE),
         VIC_SEX=as.factor(VIC_SEX)) %>%
  mutate(YEAR=year(OCCUR_DATE)) %>%
  select(YEAR,OCCUR_DATE,BORO,PRECINCT,STATISTICAL_MURDER_FLAG,
         VIC_AGE_GROUP, VIC_SEX,VIC_RACE)
```

### Analyze & tidy up NYPD data
```{r analyze_data}
## see murder data by year & disaggregated by location, sex, race
NYPD_TOTAL_MURDER <- NYPD_data %>%
  group_by(YEAR, BORO, VIC_SEX, VIC_RACE) %>%
  summarise(total_murder = sum(STATISTICAL_MURDER_FLAG))%>%
  ungroup()

## total murder by year, sex, race
NYPD_MURDER_RACE <- NYPD_TOTAL_MURDER %>%
  group_by(YEAR, VIC_SEX, VIC_RACE) %>%
  summarise(MURDER_RACE = sum(total_murder)) %>%
  ungroup
# tidy up murder by race data in columns instead of rows
# change race = "XXX" to whichever race you would like to see
race <- "BLACK"
MURDER_BY_RACE <- NYPD_MURDER_RACE %>%
  filter(VIC_RACE == race) %>%
  select(YEAR, VIC_SEX, MURDER_RACE) %>%
  pivot_wider(names_from = VIC_SEX, 
              values_from = MURDER_RACE, 
              values_fill = 0) %>%
  mutate(MURDER_SUM = rowSums(.[2:4]))

## total murder by year, location, sex
NYPD_MURDER_SEX <- NYPD_TOTAL_MURDER %>%
  group_by(YEAR, BORO, VIC_SEX) %>%
  summarise(MURDER_SEX = sum(total_murder)) %>%
  ungroup
# tidy up to see murder by boro data over the years
MURDER_BY_BORO <- NYPD_MURDER_SEX %>%
  pivot_wider(names_from = VIC_SEX, 
              values_from = MURDER_SEX, 
              values_fill = 0) %>%
  mutate(MURDER_SUM = rowSums(.[3:5])) %>%
  select(BORO,YEAR,MURDER_SUM)
MURDER_BY_BORO_CLEAN <- MURDER_BY_BORO %>%
  pivot_wider(names_from = YEAR,
              values_from = MURDER_SUM,
              values_fill = 0)
```

### Visualize & analyze NYPD data

Brief points of analysis based on visualization:

* Based on the first visualization, it can be seen that people in the race category "black" makes up the most murder victims over the years from 2006 to 2022.
* Looking further into the profiles of murder victims in the race category "black", it can be seen that a majority of these victims are male.
* Historically, Brooklyn has the highest number of murder victims but Bronx has recorded higher murder numbers in the past couple of years.
* Murder victims who are male and black consistently recorded the highest numbers in all boroughs over the years. 

```{r visualize_data}
## visualize murder victim data by race in each year
NYPD_MURDER_RACE %>%
  ggplot(aes(x=YEAR, y=MURDER_RACE, fill=VIC_RACE)) +
  geom_bar(position="stack", stat="identity") +
  theme(legend.position = "bottom") +
  labs(title=str_c("MURDER BY RACE"), y=NULL)

## visualize murder victim data by sex in each race over the years
# change race = "XXX" to whichever race you would like to see
race <- "BLACK"
NYPD_MURDER_RACE %>%
  filter(VIC_RACE == race) %>%
  ggplot(aes(x=YEAR, y=MURDER_RACE, fill=VIC_SEX)) +
  geom_bar(position="stack", stat="identity") +
  theme(legend.position = "bottom") +
  labs(title=str_c("MURDER BY SEX (", race, "), 2006-2022"), y=NULL)

## visualize murder victim data in each borough over the years
MURDER_BY_BORO %>%
  ggplot(aes(x=YEAR, y=MURDER_SUM, fill=BORO)) +
  geom_bar(position="dodge", stat="identity") +
  labs(title=str_c("MURDER IN BORO"), y=NULL)

## visualize murder victim data by sex in each borough over the years
# change boro = "XXX" to whichever boro you would like to see
boro = "BRONX"
NYPD_MURDER_SEX %>%
  filter(BORO == boro) %>%
  ggplot(aes(x=YEAR, y=MURDER_SEX, fill=VIC_SEX)) +
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label=MURDER_SEX), size=2, 
            position="stack",vjust=1) +
  theme(legend.position = "bottom") +
  labs(title=str_c("MURDER IN ", boro), y=NULL)

## visualize murder victim data in all boroughs by sex 
NYPD_MURDER_SEX %>%
  ggplot(aes(x=YEAR, y=MURDER_SEX, fill=VIC_SEX)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(~BORO) +
  theme(axis.title = element_blank()) +
  labs(title=str_c("MURDER IN BORO BY SEX"), y=NULL)
```

### Possible sources of bias & bias mitigation
1. Bias in data
  + possible bias in information gathered about perpetrators of murder as the data only include those who have been identified
  + mitigation measure: chose to analyze victim data only 
2. Bias in results of analysis
  + possible negative bias towards boroughs with higher reported numbers of murder victims
  + mitigation measure: possible next step to gather additional data to further ascertain if higher reported numbers of murder victims is a good indication of the security situation in that area

### Conclusion
People who are usually identified as "male" and "black" are in the most vulnerable group of potentially being a murder victim. Among the areas that murders are most recorded are Brooklyn and Bronx. 

It needs to be noted that the analysis and conclusion are a simple reflection of the data provided and does not imply any causation nor relation between any of the variables presented.

### Session info
```{r}
sessionInfo()
```